---
author: "Joe Colgan"
title: "**Transcriptomic analysis of fat bodies of control and inseminated _Bombus terrestris_ workers and queens**"
output:
  pdf_document: default
  html_document: default
fig_width: 4
fig_height: 4
fontsize: 20pt
---

## Introduction
The purpose of the present analysis is to examine similarities and differences in terms of gene expression associated with the fat body of _Bombus terrestris_ workers and queens. For the present study, control (unmated), and artificially inseminated bumblebees were collected and fat bodies removed at three different time points (two, four and eight days post-treatment). The study also
included treatment groups to examine changes in gene expression due to artificial insemination alone as well as the dilutant used during the insemination step.  

## Samples:  
The 'fat_body' dataset consists:  
**Queen samples:** n = 26  
- Control (n = 12; stage 1 n = 4, stage II n = 4, and stage IV n = 4)
- Insemination (n = 14; stage 1 n = 6, stage II n = 4, and stage IV n = 4)  

## Methods  
Cleaned sequences were quality assessed using FastQC ensuring high base quality and 
low adaptor contamination. Cleaned sequences were aligned against the latest 
_Bombus terrestris_ reference genome assembly using [STAR](https://github.com/alexdobin/STAR) with '--quantMode GeneCounts' parameter resulting in the generation of gene-level counts. Such counts were analysed with 
 [DESeq2](https://bioconductor.org/packages/release/bioc/html/DESeq2.html). Overlap in terms of genes expressed were analysed using upSet. Counts were transformed using variance stabilising transformation and principal component analyses performed using these values. Prior to differential expression analysis, distance matrices were calculated based on Euclidean distance to examine similarities and differences in expression profiles within and across groups. Differential expression analyses were performed using likelihood ratio tests with multiple correction performed using the Benjamini-Hochberg method (adjusted p value < 0.05) to determine global differences between castes, treatment groups and sampling stage (days post treatment). Interaction effects between caste and treatment group were also assessed using a second model. Rank-based Gene Ontology enrichment analysis for differentially expressed genes was performed using a Kolgomorov-Smirnov test implemented in topGO using the 'classic' algorithm and a node size of 20. The 'classic' algorithm was used as opposed to the 'weight01' algorithm to allow for more direct comparisons between sets of terms associated with differentially expressed genes.    
 
## Results  
The first aspect of the analysis involved examining outliers.  
First, the number of expressed genes was compared across individuals using an upset plot (Fig. 1).  
This revealed **8650** genes expressed in all samples (approximately 78% of all genes (n = 11062) expressed in the fat body).   
Second, density plots were generated for the percentage of expressed genes per sample for:
- Treatment (Fig. 2A)
- Stage (Fig. 2B)  
There was a significant difference  in terms of the mean percentage of expressed genes between stage (ANOVA; p = 0.0001) but not treatment (ANOVA; p = 0.06).  

```{r, echo = FALSE, message = FALSE, results = 'hide', warning=FALSE}
## Step One: Load libraries - if not present, install from scratch.
## Similarly, create output directory (if does not exist already)

# Load libraries; install from scratch if needed
libraries <- c("readr",
               "DESeq2",
               "ggplot2",
               "ggpubr",
               "purrr",
               "UpSetR",
               "hash",
               "reshape2",
               "eulerr",
               "knitr",
               "RUVSeq",
               "PoiClaClu",
               "genefilter",
               "pheatmap",
               "RColorBrewer",
               "WGCNA")
for (lib in libraries) {
    if (require(package = lib, character.only = TRUE)) {
        print("Successful")
    } else {
        print("Installing")
        source("https://bioconductor.org/biocLite.R")
        library(lib, character.only = TRUE )
    }
}
dir.create("results")
opts_chunk$set(fig.width = 12, fig.height = 8)
```

```{r, echo = FALSE, message = FALSE, results = 'hide', warning=FALSE}
## Set paths to folders containing output files from STAR
paths <- list()
paths$sample_batch_info <- "data/sample_information.txt"
paths$htseq_output   <- "input/fat_body/"
# Set relative paths:
paths$htseq_files_relative <- grep(x = list.files(paths$htseq_output,
                                                     recursive = TRUE),
                                      pattern = ".tab",
                                      value   = TRUE)
paths$htseq_files <- file.path(paths$htseq_output,
                               paths$htseq_files_relative)
# Automatically extract file names from kallisto output for colnames
names(paths$htseq_files) <- gsub(paths$htseq_files_relative,
                                    pattern = "/.*",
                                    replacement = "")
for (filenumber in 1:length(paths$htseq_files)) {
  current_name <- names(paths$htseq_files)[filenumber]
  current_file <- paths$htseq_files[filenumber]
  if (FALSE == grepl(pattern = current_name, x = current_file)) {
    kill("we have a problem - names and filenames dont match up")
  }
}
```

```{r, echo = FALSE, message = FALSE, results = 'hide', warning=FALSE}
# Step Three: Put sample names and treatments into a samples dataframe for DESeqDataSetFromTximport

#Differential expression analysis and data exploration was performed using [DESeq2](https://bioconductor.org/packages/release/bioc/html/DESeq2.html), which allows for identfication of genes significantly differentially expressed, as well as examination of substructure and clustering through the implementation of principal component analysis on normalised gene-level counts.
directory <- "./input/fat_body//"

## Extract sample names and put into df for tximport
samples     <- data.frame(treatment = names(paths$htseq_files))
## Create a list of samples:
sample_files <- grep(".tab",
                    list.files(directory),
                    value = TRUE)

## Read in sample information for room and batch:
samples_info <- read.table(file = paths$sample_batch_info,
                                  header = FALSE,
                                  col.names = c("sample_name",
                                                "caste",
                                                "tissue",
                                                "stage",
                                                "treatment_group"),
                                  row.names = 1)

## Subset samples related to tissue of interest:
samples_info <- subset(x = samples_info,
                       tissue == "Fat_body" &
                         caste == "Queen")

samples_info$sample_name <- row.names(samples_info)
samples_info$file_name   <- row.names(samples_info)

samples_info       <- samples_info[, c(5,6, 1:4)]
samples_info$stage <- factor(samples_info$stage)
```

```{r, echo = FALSE, message = FALSE, results = 'hide', warning=FALSE}
#Step Three: Explore gene-level accounts across samples to identify potential outliers.  

## Build DESeq2 dataset:
dds_htseq <- DESeqDataSetFromHTSeqCount(sampleTable = samples_info,
                                       directory = directory,
                                       design = ~ stage +
                                                  treatment_group)

## Extract and export raw gene-level counts:
counts_df <- as.data.frame(counts(dds_htseq))
counts_df <- counts_df[order(row.names(counts_df)), ]
colnames(counts_df) <- gsub(pattern = ".ReadsPerGene.out.tab",
                            replacement = "",
                            colnames(counts_df))

## Save:
dir.create(path = "results/fb_queen",
           recursive = TRUE)
write.table(x = counts_df,
            file = "results/fb_queen/qfat_body_gene_level_counts.txt",
            row.names = TRUE,
            col.names = TRUE,
            quote = FALSE,
            sep = "\t")
keep <- rowSums(counts(dds_htseq)) >= 10
dds_htseq <- dds_htseq[keep, ]

## Extract gene-level counts:
counts <- as.data.frame(counts(dds_htseq))
dim(counts)

## Create a list of genes per sample with at least five mapped reads:
new_list <- list()
for (name in 1:ncol(counts)){
  new_name <- paste("B", name, sep = "")
  new_list[[new_name]] <- row.names(subset(x = counts,
                                        counts[name] > 5))
}

## Plot upSet plot and save as a PDF:
pdf(file = "results/fb_queen/queen_fat_body_upset_plot.pdf",
    height = 12,
    width = 8) # or other device
UpSetR::upset(fromList(new_list), 
                  nsets = 40,
                  main.bar.color = "blue",
                  text.scale = 1.1,
                  nintersects = 30, 
                  order.by = "freq",
                  mainbar.y.label = "Number of shared genes", 
                  sets.x.label = "Genes expressed")
dev.off()
```

```{r figure1, ref.label = "figure1", fig.width = 8, fig.height = 8, echo = FALSE, message=FALSE}
UpSetR::upset(fromList(new_list), 
                  nsets = 40,
                  main.bar.color = "blue",
                  text.scale = 1.1,
                  nintersects = 30, 
                  order.by = "freq",
                  mainbar.y.label = "Number of shared genes", 
                  sets.x.label = "Genes expressed")
```

_**Fig. 1.** Upset plot displaying overlap in gene expression across all samples. For each sample, the number of expressed genes, as well as the overlap with other samples, are provided._  

```{r, echo = FALSE, message = FALSE, results = 'hide', warning=FALSE}
## Across all samples, the total number of genes detected:
total_genes <- length(unique(sort(unlist(new_list))))

percentages <- list()
for (name in names(new_list)){
  percentages[[name]] <- length(unique(sort(unlist(new_list[[name]])))) / total_genes
}

## Examine percentage of expressed genes across:
## - Treatment
## - Stage
percentages_df <- as.data.frame(t(as.data.frame(percentages)))
summary(percentages_df)

samples_info$short_name <- gsub(pattern = ".ReadsPerGene.out.tab",
                                replacement = "",
                                samples_info$sample_name)

row.names(percentages_df) <- gsub(pattern = ".ReadsPerGene.out.tab",
                                replacement = "",
                                samples_info$sample_name)

percentages_df$treatment <- samples_info[match(row.names(percentages_df),
                                               samples_info$short_name),]$treatment_group

percentages_df$stage <- samples_info[match(row.names(percentages_df),
                                           samples_info$short_name),]$stage

colnames(percentages_df)[1] <- "percentage"

density_treatment_plot <- ggplot(data = percentages_df,
       aes(x = percentage,
           colour = treatment)) +
  xlab(label = "Percentage of expressed genes") +
  geom_density() +
  theme_bw() +
  theme(axis.title = element_text(face = "bold",
                                  size = 10))

treatment_aov <- aov(percentage ~ treatment,
               data = percentages_df)
summary(treatment_aov)

density_stage_plot <- ggplot(data = percentages_df,
       aes(x = percentage,
           colour = stage)) +
  xlab(label = "Percentage of expressed genes") +
  geom_density() +
  theme_bw() +
  theme(axis.title = element_text(face = "bold",
                                  size = 10))

stage_aov <- aov(percentage ~ stage,
               data = percentages_df)
summary(stage_aov)
```


```{r figure2, ref.label = "figure2", fig.width = 8, fig.height = 5, echo = FALSE, message=FALSE}
## Generate a multi-panel plot:
ggarrange(density_treatment_plot,
          density_stage_plot,
          nrow = 1,
          ncol = 2,
          legend = "none",
          align = "hv",
          labels = c("A",
                     "B"))
```

_**Fig. 2.** A) treatment (red = "control", blue = "treatment"); and B) stage of ovarian development (blue = "two days", green = "four days", red = "eight days")._  

\newpage  

## Sample clustering.  
To examine general similarities and differences within and between group, a distance matrix based
on variance stabilising transformed (VST) counts was performed. This did highlight one sample "37" as being quite different from the rest in terms of expression profiles.  

```{r, echo = FALSE, message = FALSE, results = 'hide', warning=FALSE}
## Create a DESeq2 object:
deseq_object  <- DESeq(dds_htseq,
                       test = "LRT",
                       reduced = ~stage)  

## Transform gene-level counts using variance-stablising transformtion:
vsd <- vst(deseq_object,
             blind = FALSE)

## Save object:
saveRDS(object = vsd,
        file = "results/fb_queen/vsd_queen_fat_body_treatment.rds")

sampleDists <- dist(t(assay(vsd)))

sampleDistMatrix <- as.matrix( sampleDists )
rownames(sampleDistMatrix) <- paste( vsd$caste,
                                     vsd$treatment_group,
                                     vsd$stage,
                                     sep = " - " )
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
```

```{r figure3, ref.label = "figure3", fig.width = 8, fig.height = 6, echo = FALSE, message=FALSE}
## Rename rownames:
rownames(sampleDistMatrix) <- gsub(pattern = "Insemination_group",
                                        replacement = "AI_group_1",
                                        row.names(sampleDistMatrix))
pdf(file = "results/fb_queen/fat_body_queen_filtered_pheatmap.pdf",
    height = 8,
    width = 8)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)
dev.off()
```

_**Fig. 3.** Heatmap displaying distance matrix of genes expressed within and across samples._   

```{r, echo = FALSE, message = FALSE, results = 'hide', warning=FALSE}
#poisd <- PoissonDistance(t(counts(dds_htseq)))

#samplePoisDistMatrix <- as.matrix(poisd$dd)
#rownames(samplePoisDistMatrix) <- paste(dds_htseq$caste,
#                                        dds_htseq$treatment_group,
#                                        sep=" - " )
#colnames(samplePoisDistMatrix) <- NULL
```

```{r figure4, ref.label = "figure4", fig.width = 8, fig.height = 5, echo = FALSE, message=FALSE}
#pheatmap(samplePoisDistMatrix,
#         clustering_distance_rows = poisd$dd,
#         clustering_distance_cols = poisd$dd)
```

\newpage  

## Principal component analysis.  

To examine further similarities in terms of expression profiles across samples, a principal component analysis using VST gene values was performed. The first principal component, which explains 55% of the variance in the dataset identifies the possibility of outliers with no clear difference explained by the first two components that separate caste. There was evidence that one of the treatments may be driving the pattern (Fig. 5). It was decided to rerun the PCA with the four samples separated by PC1 removed to provide more resolution on differences between samples    
```{r, echo = FALSE, message = FALSE, results = 'hide', warning=FALSE}
## Here set the two principal components that you want to compare:
## It can any combination between 1 and 22:
#first_pc <- 2
#second_pc <- 3
## Create function for plotting:
plotPCA.san <- function(object,
                        first_pc = first_pca,
                        second_pc = second_pca,
                        intgroup = "condition",
                        ntop = 1000,
                        returnData = FALSE) {
  rv <- rowVars(assay(object))
  select <- order(rv, decreasing = TRUE)[seq_len(min(ntop,
                                                     length(rv)))]
  pca <- prcomp(t(assay(object)[select, ]))
  #return(pca)
  percentVar <- pca$sdev ^ 2 / sum(pca$sdev ^ 2)
  if (!all(intgroup %in% names(colData(object)))) {
    stop("the argument 'intgroup' should specify columns of colData(dds)")
  }
  intgroup.df <- as.data.frame(colData(object)[, intgroup, drop = FALSE])
  group <- if (length(intgroup) > 1) {
    factor(apply(intgroup.df, 1,
                 paste,
                 collapse = " : "))
  }
  else {
    colData(object)[[intgroup]]
  }
  ## Select the PCAs and percentVar that you like instead of 1 and 2
  d <- data.frame(PC1 = pca$x[, first_pc],
                  PC2 = pca$x[, second_pc],
                  group = group, 
                  intgroup.df,
                  name = colData(vsd)[,1])
  if (returnData) {
    attr(d, "percentVar") <- percentVar[first_pc:second_pc]
    return(d)
  }
    ggplot(data = d,
           aes_string(x = "PC1",
                      y = "PC2",
                      color = "group")) +
            geom_point(data = d,
                       aes(fill = group),
                       colour = "black",
                       pch = 21,
                       size = 3) +
            xlab(paste0("PC", first_pc, ": ",
                        round(percentVar[first_pc] * 100),
                        "% Variance")) +
            ylab(paste0("PC", second_pc, ": ",
                        round(percentVar[second_pc] * 100),
                        "% Variance")) +
            #coord_fixed() +
            #geom_text_repel(size = 3) +
            theme_bw() +
            theme(axis.title = element_text(size = 12,
                                            face = "bold"),
                  axis.text = element_text(size = 10,
                                           face = "plain"),
                  legend.title = element_text(size = 12,
                                              face = "bold"),
                  legend.text = element_text(size = 12,
                                              face = "plain"),
                  legend.position = "top")
}
```

Scatterplots displaying combinations of the first four principal components (PCs) for:  
- Treatmnent (Fig. 4)  
- Stage (Fig. 5)

```{r, echo = FALSE, message = FALSE, results = 'hide', warning=FALSE}
## Generate plot:
group_plot_pc1_2 <- plotPCA.san(vsd,
                          first_pc = 1,
                          second_pc = 2,
            intgroup = c("treatment_group"))
group_plot_pc1_2 <- group_plot_pc1_2 +
  scale_fill_discrete(name = "Treatment")

group_plot_pc1_3 <- plotPCA.san(vsd,
                          first_pc = 1,
                          second_pc = 3,
            intgroup = c("treatment_group"))
group_plot_pc1_3 <- group_plot_pc1_3 +
  scale_fill_discrete(name = "Treatment")

group_plot_pc1_4 <- plotPCA.san(vsd,
                          first_pc = 1,
                          second_pc = 4,
            intgroup = c("treatment_group"))
group_plot_pc1_4 <- group_plot_pc1_4 +
  scale_fill_discrete(name = "Treatment")
```

```{r figure5, ref.label = "figure5", fig.width = 8, fig.height = 5, echo = FALSE, message=FALSE}
ggarrange(group_plot_pc1_2,
          group_plot_pc1_3,
          group_plot_pc1_4,
          nrow = 1,
          ncol = 3,
          align = "hv",
          legend = "none")
```

_**Fig. 4.** Principal component analysis with first two principal components shown. Individuals are coloured by treatment (control = "red"; inseminated group = "blue)._  

```{r, echo = FALSE, message = FALSE, results = 'hide', warning=FALSE}
stage_plot_pc1_2 <- plotPCA.san(vsd,
                          first_pc = 1,
                          second_pc = 2,
            intgroup = c("stage"))
stage_plot_pc1_2 <- stage_plot_pc1_2 +
  scale_fill_discrete(name = "Stage")

stage_plot_pc1_3 <- plotPCA.san(vsd,
                          first_pc = 1,
                          second_pc = 3,
            intgroup = c("stage"))
stage_plot_pc1_3 <- stage_plot_pc1_3 +
  scale_fill_discrete(name = "Stage")

stage_plot_pc1_4 <- plotPCA.san(vsd,
                          first_pc = 1,
                          second_pc = 4,
            intgroup = c("stage"))
stage_plot_pc1_4 <- stage_plot_pc1_4 +
  scale_fill_discrete(name = "Stage")
```

```{r figure7, ref.label = "figure7", fig.width = 8, fig.height = 5, echo = FALSE, message=FALSE}
ggarrange(stage_plot_pc1_2,
          stage_plot_pc1_3,
          stage_plot_pc1_4,
          nrow = 1,
          ncol = 3,
          align = "hv",
          legend = "none")
```
_**Fig. 6.** Principal component analysis with first two principal components shown. Individuals are coloured by treatment (two days = "blue"; four days = "green"; eight days = "red")_    

```{r, echo = FALSE, message = FALSE, results = 'hide', warning=FALSE}
## Calculate principal components:
pca <- prcomp(t(assay(vsd)))

pca_df <- as.data.frame(pca$x)
## check the proportion of variance explained by each PC:
summary(pca)
```

_**Fig. 4.** Principal component analysis with first two principal components shown. Individuals are coloured by caste (queen = "red"; worker = "blue)._

Based on  PCs 1 and 2, there is no obvious outlier - move to next step:

Generate a combined plot:

```{r, message = FALSE}
pca_combined <- ggarrange(
          group_plot_pc1_2,
          group_plot_pc1_3,
          group_plot_pc1_4,
          stage_plot_pc1_2,
          stage_plot_pc1_3,
          stage_plot_pc1_4,
          nrow = 2,
          ncol = 3,
          align = "hv",
          labels = c("A",
                     "B"),
          legend = "none")

## Print to console:
pca_combined

## Save to file:
ggsave(filename = "results/fb_queen/queen_fatbody_filtered_pca_plot.pdf",
       height = 8,
       width = 8)
```

## Differential expression analysis:  
As a preliminary examination of differences in gene expression due to treatment,
all samples, irrespective of caste and stage, were included in a differential expression 
analysis.

**Model 1:** To examine the effect of **treatment** on brain gene expression, we constructed a _full model_ consisting of:  
gene expression = ~stage + treatment;   
and a _reduced model_ of:  
gene expression = ~stage  

From this model, a total of **3636** genes were identified as significantly differentiated expressed (adjusted _p_ value < 0.05; Fig. 7) across treatments.  

Applying a more stringent threshold to examine specific differences between treatment groups for significant differential expression based on adjusted
_p_ value and a minimum of absolute log2 fold change of 1, our model identifies **224** (Fig. 8)
significantly differentially expressed genes with the majority (n = **211**) having elevated 
expression in the brain of inseminated bees, which is significant deviation from equal outcome of expression profiles for differentially expressed genes (binomial test; *p* < 0.05).  

```{r, echo = FALSE, message = FALSE, results = 'hide', warning=FALSE}
## Check names of comparisons:
resultsNames(deseq_object)

res <- results(deseq_object)
```

```{r figure8, ref.label = "figure8", fig.width = 8, fig.height = 5, echo = FALSE, message=FALSE}
## Plot distribution of adjusted p values:
hist(res$padj[res$baseMean > 1],
     breaks = 0:20/20,
     col = "grey50",
     border = "white",
     main = "Histogram of adjusted p-values for differentially expressed genes",
     xlab = "Adjusted p value")
```
_**Fig. 7** Histogram of adjiusted p values for differentially expressed genes between treatment groups._    
```{r, echo = FALSE, message = FALSE, results = 'hide', warning=FALSE}
## Extract results:
deseq_object_results <- results(deseq_object,
                                      name = "treatment_group_Insemination_group_vs_Control")

## Extract significant phenotype genes:
deseq_object_results$absL2FC <- abs(deseq_object_results$log2FoldChange)
deseq_object_results_sig     <- subset(x = deseq_object_results,
                                           padj < 0.05)
nrow(deseq_object_results_sig)

## Convert DESeq2 result object to a dataframe:
deseq_object_results <- as.data.frame(deseq_object_results)
write.table(x = deseq_object_results,
            file = "results/fb_queen/queen_fb_degs.txt",
            quote = FALSE,
            row.names = TRUE,
            col.names = TRUE,
            sep = "\t")
deseq_object_results_sig <- as.data.frame(deseq_object_results_sig)

## Subset genes with evidence of significantly elevated expression in inseminated individuals:
deseq_object_insemination_sig_up     <- subset(x = deseq_object_results,
                                             padj < 0.05 &
                                            log2FoldChange >= 1)
nrow(deseq_object_insemination_sig_up)

## Subset genes with evidence of significantly reduced expression in inseminated individuals:
deseq_object_insemination_sig_down <- subset(x = deseq_object_results,
                                             padj < 0.05 &
                                             log2FoldChange <= -1)
nrow(deseq_object_insemination_sig_down)

## Plot volcano plot of:
## a) Log2FoldChange (gene expression) against:
## b) -log10 transformed p-value (significant threshold)
volcano_plot <- ggplot(deseq_object_results,
       aes(x = log2FoldChange,
           y = -log10(padj))) +
  xlim(-10, 10) +
  geom_point(alpha = 0.5) +
  geom_point(data = subset(deseq_object_insemination_sig_up,
                           log2FoldChange >= 1 &
                             padj < 0.05),
             aes(colour = "orange")) +
  geom_point(data = subset(deseq_object_insemination_sig_down,
                           log2FoldChange <= -1 &
                             padj < 0.05),
             aes(colour = "blue")) +
  geom_vline(xintercept = c(-1, 1),
             linetype = "dashed") +
  theme_bw() +
  theme(
    axis.title = element_text(face = "bold",
                              size = 12),
    legend.title = element_text(face = "bold",
                                size = 12),
    legend.text = element_text(face = "plain",
                                size = 12),
    legend.position = "top") +
  scale_colour_manual(values = c("blue",
                                 "orange"),
                      labels = c("Inseminated-reduced",
                                 "Inseminated-biased"),
                      name = "Status")
```

## Interaction effect between treatment and stage:  
**Model 2:** We can examine an interaction between treatment and stage using
only queen samples:

Full model: gene expression = ~treatment + stage + caste:stage 

We then construct a _reduced model_ consisting of:
gene expression = ~treatment + stage  

Analysis identified **447** genes as differentially expressed (adjusted _p_ value < 0.05) between treatment depending on stage.  

```{r, echo = FALSE, message = FALSE, results = 'hide', warning=FALSE}
## Build DESeq2 dataset:
design(deseq_object) <- ~ stage +
                          treatment_group +
                          treatment_group:stage

## Create a DESeq2 object:
deseq_object  <- DESeq(deseq_object,
                       test = "LRT",
                       reduced = ~stage +
                                  treatment_group)

## Check names of comparisons:
resultsNames(deseq_object)
## Extract results:
deseq_object_results <- results(deseq_object,
                                      name = "stageStage_II.treatment_groupInsemination_group")

## Extract significant phenotype genes:
deseq_object_results_sig     <- subset(x = deseq_object_results,
                                             padj < 0.05)
nrow(deseq_object_results_sig)

## Convert DESeq2 result object to a dataframe:
deseq_object_results <- as.data.frame(deseq_object_results)
deseq_object_results_sig <- as.data.frame(deseq_object_results_sig)
```
