---
author: "Joe Colgan"
title: "**Weight gene co-expression analysis of control and inseminated _Bombus terrestris_ workers and queens**"
output:
  pdf_document: default
  html_document: default
fig_width: 4
fig_height: 4
fontsize: 20pt
---

## Introduction
The purpose of the present analysis is to examine similarities and differences in terms of gene expression associated with the reproductive tissues (spermatheca, vagina, median oviduct) of _Bombus terrestris_ workers and queens. For the present study, control (unmated) and artificially inseminated bumblebees were collected and tissues removed at three different time points (two, four and eight days post-treatment).

The analysis provided here follows steps outlined by the authors of WGCNA - additional information can be found here:  
https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/
  
1. Load libraries:

```{r, message = FALSE}
library(WGCNA)
library(DESeq2)
library(igraph)
library(ggcorrplot)
library(ggplot2)
library(ggpubr)
```

2. Load input data:

Load data, which should include normalised gene-level counts. Here, normalisation was performed using variance-stablising transformation as performed by DESeq2.

```{r, message = FALSE}
# The following setting is important, do not omit.
options(stringsAsFactors = FALSE)

## Load experimental data:
data <- readRDS(file = "input/vsd_spermatheca_treatment.rds")
data_counts <- assay(data)
```

3. Transpose the input dataframe so each column is a gene and row is a sample.  

```{r, message=FALSE}
## We need a transposed dataframe consisting of gene per column and sample per
## row:
transposed_data <- t(data_counts)
```

4. Check data for excessive missing values and identification of outlier samples

We first check for genes and samples with too many missing values:

```{r, message = FALSE}
## Run for experimental dataset:
transposed_gsg <- goodSamplesGenes(transposed_data, verbose = 3)
transposed_gsg$allOK
```

If the last statement returns TRUE, all genes have passed the cuts. If not, we remove the offending genes and samples from the data:

```{r, message = FALSE}
## Run for experimental:
if (!transposed_gsg$allOK) {
  # Optionally, print the gene and sample names that were removed:
  if (sum(!transposed_gsg$goodGenes) > 0)
     printFlush(paste("Removing genes:",
                      paste(names(transposed_data)[!transposed_gsg$goodGenes],
                            collapse = ", ")))
  if (sum(!transposed_gsg$goodSamples) > 0)
     printFlush(paste("Removing samples:",
                  paste(rownames(transposed_data)[!transposed_gsg$goodSamples],
                            collapse = ", ")))
  # Remove the offending genes and samples from the data:
  transposed_data <- transposed_data[transposed_gsg$goodSamples,
                                    transposed_gsg$goodGenes]
}
```

5. Next we cluster the samples (in contrast to clustering genes that will come later) to see if there are any obvious outliers.

```{r, message = FALSE}
## Run for experimental samples:
exp_sample_tree <- hclust(dist(transposed_data),
                         method = "average")
# Plot the sample tree: Open a graphic output window of size 12 by 9 inches
# The user should change the dimensions if the window is too large or too small.
sizeGrWindow(12, 9)
par(cex = 0.6)
par(mar = c(0, 4, 2, 0))
plot(exp_sample_tree, main = "Sample clustering to detect outliers",
     sub = "",
     xlab = "",
     cex.lab = 1.5,
     cex.axis = 1.5,
     cex.main = 2)
```

We filter out identified outlier samples

```{r, message = FALSE}
# Determine cluster under the line
clust <- cutreeStatic(exp_sampleTree,
                     cutHeight = 150,
                     minSize = 10)
table(clust)
# clust 1 contains the samples we want to keep.
keep_samples <- (clust == 1)
transposed_data_filt <- transposed_data[keep_samples, ]
n_genes <- ncol(transposed_data_filt)
n_samples <- nrow(transposed_data_filt)
```

The object transposed_data_filt now contains the expression data ready for network analysis.  

6. Loading sample information:
We now read in sample information metadata and match to the expression samples.  
For our experiment, we have three attributes of interest:
- Caste
- Treatment
- Age or stage of ovarian development depending on tissue

```{r, message = FALSE}
## Read in sample information:
samples_information <- read.table(file = "data/sample_info_for_wgcna.txt",
                                  header = FALSE)
## Remove tissue:
samples_information$V3 <- NULL
row.names(samples_information) <- samples_information$V1
samples_information$V1 <- NULL
## Update column names:
colnames(samples_information) <- c("caste",
                                   "age",
                                   "treatment")
## Remove outlier:
samples_information <- subset(samples_information,
                              row.names(samples_information) %in%
                                row.names(transposed_data_filt))
```

Before network construction and module detection, we visualise how the attributes relate to the sample dendrogram.  

```{r, message = FALSE}
## Run for experimental dataset:
# Re-cluster samples
exp_sample_tree2 <- hclust(dist(transposed_data_filt),
                         method = "average")
# Convert traits to a color representation: white means low, red means high,
# grey means missing entry
exp_trait_colors <- numbers2colors(samples_information, signed = FALSE)

exp_sample_tree2$labels <- gsub(pattern = ".ReadsPerGene.out.tab",
                               replacement = "",
                               exp_sample_tree2$labels)

# Plot the sample dendrogram and the colors underneath:
pdf(file = "results/sample_dendrogram_trait_heatmap.pdf",
    height = 6,
    width = 8)
plotDendroAndColors(exp_sample_tree2,
                    exp_trait_colors,
                  groupLabels = gsub(pattern = ".ReadsPerGene.out.tab",
                  replacement = "",
                  names(samples_information)),
                  main = "Sample dendrogram and trait heatmap")
dev.off()

saveRDS(object = exp_sample_tree2,
        file = "results/exp_sample_tree2_spermatheca.rds")
saveRDS(object = exp_trait_colors,
        file = "results/exp_trait_colors_spermatheca.rds")
```

7 Step-by-step network construction and module detection:  
7.1 Choosing the soft-thresholding power: analysis of network topology
Constructing a weighted gene network entails the choice of the soft thresholding power β to which co-expression similarity is raised to calculate adjacency. 

```{r, message = FALSE}
## For the experimental set:
# Choose a set of soft-thresholding powers
powers <- c(c(1:10), seq(from = 12,
                        to = 20,
                        by = 2))

# Call the network topology analysis function
exp_sft <- pickSoftThreshold(transposed_data_filt,
                            powerVector = powers,
                            verbose = 5)
# Plot the results:
sizeGrWindow(9, 5)
par(mfrow = c(1, 2))
cex1 <- 0.9
# Scale-free topology fit index as a function of the soft-thresholding power
plot(exp_sft$fitIndices[, 1],
     -sign(exp_sft$fitIndices[, 3]) * exp_sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit,signed R^2",
     type = "n",
     main = paste("Scale independence"))
text(exp_sft$fitIndices[, 1],
     -sign(exp_sft$fitIndices[, 3]) * exp_sft$fitIndices[, 2],
     labels = powers,
     cex = cex1,
     col = "red")
# this line corresponds to using an R^2 cut-off of h
abline(h = 0.90,
       col = "red")
# Mean connectivity as a function of the soft-thresholding power
plot(exp_sft$fitIndices[, 1],
     exp_sft$fitIndices[, 5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type = "n",
     main = paste("Mean connectivity"))
text(exp_sft$fitIndices[, 1],
     exp_sft$fitIndices[, 5],
     labels = powers,
     cex = cex1,
     col = "red")
```

7.2 Co-expression similarity and adjacency
We now calculate the adjencies, using the soft thresholding power 6:

```{r, message = FALSE}
## For experimental dataset:
soft_power <- 6
exp_adjacency <- adjacency(transposed_data_filt,
                          power = soft_power)
```

7.3 Topological Overlap Matrix (TOM)
To minimize effects of noise and spurious associations, we transform the adjacency into a Topological Overlap Matrix (TOM), and calculate the corresponding dissimilarity:

```{r, message = FALSE}
## Experimental dataset:
# Turn adjacency into topological overlap
exp_tom <- TOMsimilarity(exp_adjacency)
diss_exp_tom <- 1 - exp_tom
```

7.4 Clustering using TOM
We now use hierarchical clustering to produce a hierarchical clustering tree (dendrogram) of genes:  

```{r, message = FALSE}
## Experimental data:
exp_gene_tree <- hclust(as.dist(diss_exp_tom),
                      method = "average")
# Plot the resulting clustering tree (dendrogram)
sizeGrWindow(12, 9)
plot(exp_gene_tree,
     xlab = "",
     sub = "",
     main = "Gene clustering on TOM-based dissimilarity",
     labels = FALSE,
     hang = 0.04)
```

In the clustering tree (dendrogram), each leaf, that is a short vertical line, corresponds to a gene. Branches of the dendrogram group together densely interconnected, highly co-expressed genes. Module identification amounts to the identification of individual branches (”cutting the branches off the dendrogram”).

```{r, message = FALSE}
## Experimental set:
# We like large modules, so we set the minimum module size relatively high:
min_module_size <- 30
# Module identification using dynamic tree cut:
exp_dynamic_mods <- cutreeDynamic(dendro = exp_gene_tree,
                                distM = diss_exp_tom,
              deepSplit = 2,
              pamRespectsDendro = FALSE,
              minClusterSize = min_module_size)
table(exp_dynamic_mods)
```

We now plot the module assignment under the gene dendrogram:

```{r, message=FALSE}
## Experimental data:
# Convert numeric lables into colors
exp_dynamic_colors <- labels2colors(exp_dynamic_mods)
table(exp_dynamic_colors)
# Plot the dendrogram and colors underneath
sizeGrWindow(8, 6)
plotDendroAndColors(exp_gene_tree,
                    exp_dynamic_colors,
                    "Dynamic Tree Cut",
                    dendroLabels = FALSE,
                    hang = 0.03,
                    addGuide = TRUE,
                    guideHang = 0.05,
                    main = "Gene dendrogram and module colors")
```

7.5 Merging of modules whose expression profiles are very similar
The Dynamic Tree Cut may identify modules whose expression profiles are very similar. It may be prudent to merge such modules since their genes are highly co-expressed. To quantify co-expression similarity of entire modules, we calculate their eigengenes and cluster them on their correlation:   

```{r, message = FALSE}
## Experimental dataset:
# Calculate eigengenes:
exp_me_list <- moduleEigengenes(transposed_data_filt,
                              colors = exp_dynamic_colors)
exp_mes <- exp_me_list$eigengenes
# Calculate dissimilarity of module eigengenes
exp_me_diss <- 1 - cor(exp_mes)
# Cluster module eigengenes
exp_me_tree <- hclust(as.dist(exp_me_diss),
                    method = "average")
# Plot the result
sizeGrWindow(7, 6)
plot(exp_me_tree,
     main = "Clustering of module eigengenes",
     xlab = "",
     sub = "")
```

We choose a height cut of 0.25, corresponding to correlation of 0.75, to merge:

```{r, message = FALSE}
## for the experimental dataset:
exp_me_diss_thres <- 0.25
# Call an automatic merging function
exp_merge <- mergeCloseModules(transposed_data_filt,
                              exp_dynamic_colors,
                              cutHeight = exp_me_diss_thres,
                              verbose = 3)
# The merged module colors
exp_merged_colors <- exp_merge$colors
# Eigengenes of the new merged modules:
exp_merged_mes <- exp_merge$newMEs
```

To see what the merging did to our module colors, we plot the gene dendrogram again, with the original and merged module colors underneath.

```{r, message = FALSE}
sizeGrWindow(12, 9)
plotDendroAndColors(exp_gene_tree,
                    cbind(exp_dynamic_colors,
                          exp_merged_colors),
                  c("Dynamic Tree Cut",
                    "Merged dynamic"),
                  dendroLabels = FALSE,
                  hang = 0.03,
                  addGuide = TRUE,
                  guideHang = 0.05)
```

In the subsequent analysis, we will use the merged module colors in mergedColors. We save the relevant variables for use in subsequent parts of the tutorial:

```{r, message = FALSE}
## Experimental dataset:
# Rename to moduleColors
exp_module_colors <- exp_merged_colors
# Construct numerical labels corresponding to the colors
exp_color_order <- c("grey", standardColors(50))
exp_module_labels <- match(exp_module_colors,
                         exp_color_order) - 1
exp_mes <- exp_merged_mes
dir.create(path = "results")
# Save module colors and labels for use in subsequent parts
save(exp_mes,
     exp_module_labels,
     exp_module_colors,
     exp_gene_tree,
     file = "results/spermatheca-02-networkConstruction-stepByStep.RData")
```

8. Relating modules to attributes:
8.1 Quantifying module–trait associations
In this analysis we would like to identify modules that are significantly associated with the measured traits. Since we already have a summary profile (eigengene) for each module, we simply correlate eigengenes with external traits and look for the most significant associations:

```{r, message = FALSE}
# Define numbers of genes and samples
exp_n_genes <- ncol(transposed_data_filt)
exp_n_samples <- nrow(transposed_data_filt)
# Recalculate MEs with color labels
exp_mes0 <- moduleEigengenes(transposed_data_filt,
                            exp_module_colors)$eigengenes
exp_mes <- orderMEs(exp_mes0)
exp_module_trait_cor <- cor(exp_mes,
                         samples_information,
                         use = "p")
exp_module_trait_pvalue <- corPvalueStudent(exp_module_trait_cor,
                                            exp_n_samples)
```

We colour code each association by the correlation value:  

```{r, message = FALSE}
## Experimental dataset:
sizeGrWindow(10, 6)
# Will display correlations and their p-values
exp_text_matrix <- paste(signif(exp_module_trait_cor,
                               2),
                        "\n(",
                        signif(exp_module_trait_pvalue,
                               1), ")",
                        sep = "")
dim(exp_text_matrix) <- dim(exp_module_trait_cor)
# Display the correlation values within a heatmap plot:
pdf(file = "results/module-trait_relationship_heatmap.pdf",
    height = 6,
    width = 8)
par(mar = c(6, 8.5, 3, 3))
labeledHeatmap(Matrix = exp_module_trait_cor,
             xLabels = names(samples_information),
             yLabels = names(exp_mes),
             ySymbols = names(exp_mes),
             colorLabels = FALSE,
             colors = blueWhiteRed(50),
             textMatrix = exp_text_matrix,
             setStdMargins = FALSE,
             cex.text = 0.5,
             zlim = c(-1, 1),
             main = paste("Module-trait relationships"))
dev.off()

# Save module colors and labels for use in subsequent parts
save(exp_MEs,
     exp_module_trait_cor,
     samples_information,
     exp_text_matrix,
     file = "results/spermatheca-03-input_for_labelled_heatmap.RData")
```

8.2 Gene relationship to attribute and important modules: Gene Significance and Module Membership
We quantify associations of individual genes with our trait of interest (treatment) by defining Gene Significance GS as (the absolute value of) the correlation between the gene and the trait. For each module, we also define a quantitative measure of module membership MM as the correlation of the module eigengene and the gene expression profile. This allows us to quantify the similarity of all genes on the array to every module.

```{r, message = FALSE}
# Define variable treatment containing the treatment column of samples
# information
treatment <- as.data.frame(samples_information$treatment)
names(treatment) <- "treatment"
# names (colors) of the modules
exp_mod_names <- substring(names(exp_mes), 3)
exp_gene_module_membership <- as.data.frame(cor(transposed_data_filt,
                                         exp_mes,
                                         use = "p"))
exp_mm_pvalue <- as.data.frame(corPvalueStudent(as.matrix(exp_gene_module_membership),
                                          exp_n_samples))

names(exp_gene_module_membership) <- paste("MM",
                                        exp_mod_names,
                                        sep = "")
names(exp_mm_pvalue) <- paste("p.MM",
                            exp_mod_names,
                            sep = "")

exp_gene_trait_significance <- as.data.frame(cor(transposed_data_filt,
                                          treatment,
                                          use = "p"))
exp_gs_pvalue <- as.data.frame(corPvalueStudent(as.matrix(exp_gene_trait_significance),
                                          exp_n_samples))
names(exp_gene_trait_significance) <- paste("GS.",
                                         names(treatment),
                                         sep = "")
names(exp_gs_pvalue) <- paste("p.GS.",
                            names(treatment),
                            sep = "")
```

8.3 Intramodular analysis: identifying genes with high GS and MM
Using the GS and MM measures, we can identify genes that have a high significance for treatment as well as high module membership in interesting modules:

```{r, message = FALSE}
exp_module <- "yellow"
exp_column <- match(exp_module,
                   exp_mod_names)
exp_module_genes <- exp_module_colors == exp_module
sizeGrWindow(7, 7)
par(mfrow = c(1, 1))
verboseScatterplot(abs(exp_gene_module_membership[exp_module_genes, exp_column]),
                   abs(exp_gene_trait_significance[exp_module_genes, 1]),
                 xlab = paste("Module Membership in", exp_module, "module"),
                 ylab = "Gene significance for treatment",
                 main = paste("Module membership vs. gene significance\n"),
                 cex.main = 1.2,
                 cex.lab = 1.2,
                 cex.axis = 1.2,
                 col = exp_module)

# Save module colors and labels for use in subsequent parts
save(exp_gene_module_membership,
     exp_module_genes,
     exp_module,
     exp_column,
     exp_gene_trait_significance,
     file = "results/spermatheca-04-input_for_membership_scatterplot_yellow.RData")

## Generate a combined plot:
pdf(file = "results/module-trait_relationship_heatmap.pdf",
    height = 6,
    width = 12)
par(mar = c(6, 8.5, 3, 3))
labeledHeatmap(Matrix = exp_module_trait_cor[1:5, ],
             xLabels = names(samples_information),
             yLabels = names(exp_mes)[1:5],
             ySymbols = names(exp_mes)[1:5],
             colorLabels = FALSE,
             colors = blueWhiteRed(50),
             textMatrix = exp_text_matrix[1:5, ],
             setStdMargins = FALSE,
             cex.text = 1,
             zlim = c(-1, 1),
             main = paste(""))
dev.off()

verboseScatterplot(abs(exp_gene_module_membership[exp_module_genes, exp_column]),
                   abs(exp_gene_trait_significance[exp_module_genes, 1]),
                 xlab = paste("Module Membership in", exp_module, "module"),
                 ylab = "Gene significance for treatment",
                 main = paste("Module membership vs. gene significance\n"),
                 cex.main = 1.2,
                 cex.lab = 1.2,
                 cex.axis = 1.2,
                 col = exp_module)
```

9. Extract and save the name of genes associated with a particular module:

````{r, message = FALSE}
transposed_data_filt <- as.data.frame(transposed_data_filt)
names(transposed_data_filt)
red_module_genes <- names(transposed_data_filt)[exp_module_colors == "red"]
length(red_module_genes)
write(x = red_module_genes,
      file = "results/red_module_genes.txt")

yellow_module_genes <- names(transposed_data_filt)[exp_module_colors == "yellow"]
length(yellow_module_genes)
write(x = yellow_module_genes,
      file = "results/yellow_module_genes.txt")

pink_module_genes <- names(transposed_data_filt)[exp_module_colors == "pink"]
length(pink_module_genes)
write(x = pink_module_genes,
      file = "results/pink_module_genes.txt")

turquoise_module_genes <- names(transposed_data_filt)[exp_module_colors == "turquoise"]
length(turquoise_module_genes)
write(x = turquoise_module_genes,
      file = "results/turquoise_module_genes.txt")

grey_module_genes <- names(transposed_data_filt)[exp_module_colors == "grey"]
length(grey_module_genes)
write(x = grey_module_genes,
      file = "results/grey_module_genes.txt")

black_module_genes <- names(transposed_data_filt)[exp_module_colors == "black"]
length(black_module_genes)
write(x = black_module_genes,
      file = "results/black_module_genes.txt")

## Choose top hub for each module:
chooseTopHubInEachModule(transposed_data_filt,
                         colorh = exp_module_colors)
```

Examination of enriched Gene Ontology (GO) terms for genes assigned to 
modules of interest:

GO term enrichment analysis is performed using scripts provided on the project
Github. Here, we import the results of those analyses and plot. 

Read back in enriched Gene Ontology terms for:
- Genes belonging to the red module
- Genes belonging to the yellow module

```{r, message = FALSE}
## Read in Gene Ontology terms enriched for resident genes:
go_term_dir <- "../2022_gene_ontology_analysis/run_reps/results/bterrestris_dmel"
## Load ALH output:
red_bp_terms <- read.table(file = paste(go_term_dir,
                                        "/output_20_red_module_fisher_weight01",
                                        "/BP_sig.tmp",
                                              sep = ""),
                           header = TRUE)
red_bp_terms$log10 <- -log10(red_bp_terms$weight_fisher)
red_bp_terms$category <- "Red module genes"

## Subset by significance:
yellow_bp_terms <- read.table(file = paste(go_term_dir,
                                          "/output_20_yellow_module_fisher_weight01",
                                          "/BP_sig.tmp",
                                              sep = ""),
                           header = TRUE)
yellow_bp_terms$log10 <- -log10(yellow_bp_terms$weight_fisher)
yellow_bp_terms$category <- "Yellow module genes"

## Generate plot:
combined_df <- as.data.frame(rbind(red_bp_terms,
                                   yellow_bp_terms))
combined_df_unique <- subset(combined_df,
                             !duplicated(combined_df$GO.ID))

## Reannotate:
combined_df_unique$Term <- gsub(pattern = "_",
                                replacement = " ",
                                combined_df_unique$Term)

## Create a column containing amend terms:
combined_df_unique$amended_terms <- paste(combined_df_unique$Term,
                                          " ",
                                          "(",
                                          combined_df_unique$Annotated,
                                          ")",
                                          sep = "")

## Relevel category factors:
combined_df_unique$category <- factor(combined_df_unique$category,
                                      levels = c("Red module genes",
                                                 "Yellow module genes"))

## Generate plot:
plot <- ggbarplot(combined_df_unique,
                  x = "amended_terms",
                  y = "log10",
                  position = position_dodge(0.1),
                  fill = "category",
                  color = NULL,
                  palette = "jco",
                  sort.val = "desc",
                  sort.by.groups = TRUE,
                  ylab = "-log10(p)",
                  xlab = "Enriched Gene Ontology term",
                  legend.title = "Gene categories",
                  x.text.angle = 90,
                  lab.col = "black",
                  lab.size = 4,
                  lab.vjust = 0.5,
                  lab.hjust = 1,
                  legend = "top",
                  rotate = FALSE,
                  ggtheme = theme_minimal())

## Make font bigger and bold:
plot <- plot +
  scale_fill_manual(values = c("red",
                               "gold")) +
                scale_y_continuous(expand = c(0, 0)) +
                theme(axis.text = element_text(size = 10),
                      axis.title.x = element_text(size = 12,
                                                  face = "bold"),
                      axis.title.y = element_text(size = 12,
                                                  face = "bold"),
                      axis.text.y = element_text(size = 10,
                                                 face = "bold"),
                      axis.text.x = element_text(size = 10),
                      legend.position = "top",
                      legend.title = element_text(size = 10,
                                                  face = "bold")) +
                expand_limits(y = 10) +
                geom_hline(yintercept = 1.301,
                           linetype = "dashed",
                           colour = "black") +
                geom_hline(yintercept = -1.301,
                           linetype = "dashed",
                           colour = "black")

## Generate a correlation plot using ggcorrplot:
correlation_plot <- ggcorrplot(exp_moduleTraitCor,
           lab = TRUE) +
  theme(axis.text = element_text(size = 10,
                                  face = "bold"),
        axis.text.x = element_text(size = 10,
                                   colour = c("turquoise",
                                              "pink",
                                              "black",
                                              "red",
                                              "gold",
                                              "darkgrey"),
                                  face = "bold"),
        legend.title = element_text(size = 10,
                                  face = "bold"))

## Generate a combined plot:
ggarrange(correlation_plot,
          plot,
          ncol = 1,
          nrow = 2,
          labels = c("a",
                     "b"),
          heights = c(1, 1.5))

## Save image:
ggsave(file = "results/combined_cor_go_term_plot.pdf",
       height = 12,
       width = 8,
       dpi = 600)
```

10. Run lint:

```{r, message = FALSE}
lintr::lint(file = "run_wgcna.Rmd")
```

